Verification of the CVM algorithm with a New Recursive Analysis Technique

### 2025-02-07 paper/sec/formalization.tex
Paper:   * Add section on formalization of negatively dependend variables.   * Minor edits.

\begin{algorithm}[h]

@@

A snippet of the formalization of the new version of the algorithm is presented in \cref{alg:cvm_formalized}.
The snippet also includes the relevant correctness statements, both unbiasedness and tail bounds. 
We use the same variables as in the informal presentation: $n$ for the maximal size of the buffer, $f$ for the fraction of elements to keep in the buffer when subsampling.
The condition \isa{{\isacartoucheopen}n\ {\isacharasterisk}{\kern0pt}\ f\ {\isasymin}\ {\isasymnat}{\isacartoucheclose}} expresses the requirement that the $nf$ must be integer.
Instead of representing the state using pairs, as we did in the informal proof, we use a datatype with the single constructor \isa{State}, which has two arguments $\chi$ and $p$, the buffer and the probability, that the stream elements are in the buffer.
Isabelle provides notation closely related to informal pseudocode, such that it is usually feasible to read a formalization without expert knowledge.
For readers not familiar with Isabelle, the following table contains brief descriptions of the syntactic elements used in the listing:

\medskip
\noindent\begin{tabular}{l p{9cm}}
\toprule
Term & Description \\
\midrule
\isa{card\ S} & Cardinality of a finite set $S$. \\
\isa{set\ xs} & For a sequence \isa{xs} the set of elements in the sequence. \\
\isa{length\ xs} & For a sequence \isa{xs} the length of the sequence. \\
\isa{real} & Type of real numbers and conversion between natural numbers into real numbers. \\
\isa{nat} & Type of natural numbers (non-negative integers). \\
\isa{bernoulli{\isacharunderscore}pmf\ p} & The probability space over the boolean values, where the probability of \isa{True} is $p$. (Bernoulli distribution.) \\
\isa{pmf{\isacharunderscore}of{\isacharunderscore}set\ S} & For a finite set $S$, the uniform probability space on $S$. (Every element of $S$ is equiprobable.) \\
\isa{map{\isacharunderscore}pmf\ f\ A} & The probability space representing the distribution of the random variable $f$ over the probability space $A$. \\
\isa{return{\isacharunderscore}pmf\ x} & The probability space of the singleton $\{x\}$. \\
\isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} & Iterate randomized algorithm $f$ over the sequence $xs$ using the initial state $a$. \\
\isa{\isasymP{\isacharparenleft}x\ in\ M{\isachardot}\ P\ x\isacharparenright} & Probability of predicate $P$ in the probability space $M$. \\
\isa{measure{\isacharunderscore}pmf{\isachardot}expectation\ M\ f} & Expectation of the random variable $f$ over the probability space $M$. \\
\bottomrule
\end{tabular}

\medskip
The formalization contains four theory files. The file CVM{\textunderscore}Preliminary definitions and lemmas, such as properties about the \isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} operation we introduced.
Of course, we do not verify \cref{alg:cvm} and \cref{alg:cvm_new} separately,
as most of the proof is identical.
The file CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, with an abstract subsampling operation, which is required to fulfill \cref{eq:subsample_condition}.
The specialization happens in the following theories CVM{\textunderscore}Original{\textunderscore}Algorithm (resp. CVM{\textunderscore}New{\textunderscore}Algorithm).
It should be noted that only the last file, for the new unbiased algorithm, dependends on the new library (\cref{sec:formalization_neg_dep}) on negatively associated random variables.
The total number of lines required for the verification of the original algorithm is 1004 lines.
This is in contrast to the original proof, for which the verification required \todo{x} lines.

>>>>

\begin{algorithm}[t]

@@

A snippet of the formalization of the new version of the algorithm is presented in \cref{alg:cvm_formalized}.
The snippet also includes the relevant correctness statements, both unbiasedness and tail bounds. 
We use the same variables as in the informal presentation: $n$ for the maximal size of the buffer, $f$ for the fraction of elements to keep in the buffer when subsampling.
The condition \isa{{\isacartoucheopen}n\ {\isacharasterisk}{\kern0pt}\ f\ {\isasymin}\ {\isasymnat}{\isacartoucheclose}} expresses the requirement that the $nf$ must be integer.
Instead of representing the state using pairs, as we did in the informal proof, we use a datatype with the single constructor \isa{State}, which has two arguments $\chi$ and $p$, the buffer and the probability, that the stream elements are in the buffer.
Isabelle provides notation closely related to informal pseudocode, such that it is usually feasible to read a formalization without expert knowledge.
For readers not familiar with Isabelle, \cref{tab:isabelle_syntax} contains brief descriptions of the syntactic elements used in the listing.
%\medskip
\begin{table}
\caption{Isabelle syntax used in \cref{alg:cvm_formalized}.}\label{tab:isabelle_syntax}
\noindent\begin{tabular}{l p{9cm}}
\toprule
Term & Description \\
\midrule
\isa{card\ S} & Cardinality of a finite set $S$. \\
\isa{set\ xs} & For a sequence \isa{xs} the set of elements in the sequence. \\
\isa{length\ xs} & For a sequence \isa{xs} the length of the sequence. \\
\isa{real} & Type of real numbers and conversion between natural numbers into real numbers. \\
\isa{nat} & Type of natural numbers (non-negative integers). \\
\isa{bernoulli{\isacharunderscore}pmf\ p} & The probability space over the boolean values, where the probability of \isa{True} is $p$. (Bernoulli distribution.) \\
\isa{pmf{\isacharunderscore}of{\isacharunderscore}set\ S} & For a finite set $S$, the uniform probability space on $S$. (Every element of $S$ is equiprobable.) \\
\isa{map{\isacharunderscore}pmf\ f\ A} & The probability space representing the distribution of the random variable $f$ over the probability space $A$. \\
\isa{return{\isacharunderscore}pmf\ x} & The probability space of the singleton $\{x\}$. \\
\isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} & Iterate randomized algorithm $f$ over the sequence $xs$ using the initial state $a$. \\
\isa{\isasymP{\isacharparenleft}x\ in\ M{\isachardot}\ P\ x\isacharparenright} & Probability of predicate $P$ in the probability space $M$. \\
\isa{measure{\isacharunderscore}pmf{\isachardot}expectation\ M\ f} & Expectation of the random variable $f$ over the probability space $M$. \\
\bottomrule
\end{tabular}
\end{table}
%\medskip
The formalization contains four theory files. The file CVM{\textunderscore}Preliminary definitions and lemmas, such as properties about the \isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} operation we introduced.
Of course, we do not verify \cref{alg:cvm} and \cref{alg:cvm_new} separately,
as most of the proof is identical.
The file CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, with an abstract subsampling operation, which is required to fulfill \cref{eq:subsample_condition}.
The specialization happens in the following theories CVM{\textunderscore}Original{\textunderscore}Algorithm (resp. CVM{\textunderscore}New{\textunderscore}Algorithm).
It should be noted that only the last file, for the new unbiased algorithm, dependends on the new library (\cref{sec:formalization_neg_dep}) on negatively associated random variables.
The total number of lines required for the verification of the original algorithm is 1004 lines.
This is in contrast to the original proof, for which the verification required \todo{x} lines.

### 2025-02-07 paper/sec/formalization_neg_dep.tex
Paper:   * Add section on formalization of negatively dependend variables.   * Minor edits.

\section{Formalization of the concept of negative association}\label{sec:formalization_neg_dep}
TODO

>>>>

\section{Formalization of the concept of negative association}\label{sec:formalization_neg_dep}
As mentioned in the previous section, the formalization of the new unbiased variant of the CVM algorithm, requires results from the theory of negative association.
The supplementary material contains a formalization of the concept in the directory Negative{\textunderscore}Association, which we intend to submit to the AFP as a separate contribution.

The formalization follows the definitions by Joag-Dev and Proschan~\cite{joagdev1983} closely.
However, the authors did not correctly define the constrains for the possible set of test functions $f$ and $g$ in \cref{def:neg_assoc}, which does not make sense formally.
In the formalization, we assume that the functions are bounded and measurable.
On the other hand, we derive that the condition is still true, if $f, g$ are square integrable.
Or alternatively integrable, and non-negative.
This is derived using the monotone-convergence theorem.

Another deviation from the original work is that we do not require that the random variables need to be real-valued.
In the formalization any linearly ordered topological space with the Borel $\sigma$-algebra is allowed as the range space.
In this case the test functions must be monotone with respect ot the respective order on that space.

The formalization contains the standard Chernoff bounds for negatively associated random variables.
This includes the additive bounds by Hoeffding~\cite[Th. 1, 2]{hoeffding1963} and the multipicative bounds by Motwani and Raghavan~\cite[Th. 4.1, 4.2]{motwani1995}.
Although, these results are not needed for the proof of the CVM algorithm.

A key issue, we faced during formalization, was the fact that there are many theorems of the form if a set of functions are either simultaneously monotone or anti-monotone.
For this we introduce a notation, that allows us to abstract over the direction of relations: \isa{\isasymle\isasymge\isactrlbsub\isasymeta\isactrlesub}.
Which evaluates to the forward version of the relation $\leq$ if \isa{\isasymeta\ \isacharequal\ Fwd} and the converse: $\geq$ if \isa{\isasymeta\ \isacharequal\ Rev}.
For example the FKG inequality
\[
  \expect [f g] \geq \expect [f] \expect [g]
\]
is true, if $f$ and $g$ are both montone, or both antimontone.
On the other hand the reverse inequality is true, if $f$ is montone and $g$ is antimontone, or vice versa.
Using our parameterized relation symbol, we could state all variants in a concise manner.

### 2025-02-07 paper/sec/negdep.tex
Paper: Section on the original proof.
Paper:   * Add section on formalization of negatively dependend variables.   * Minor edits.

\section{Negative Dependence and an Unbiased CVM Algorithm\label{sec:negdep}}

Sometimes, sets of random variables possess a property called negative association.
The concept is a generalization of independence.
It has similar closure properties as independent random variables.
More interestingly, the standard Chernoff-Hoeffding bounds still apply to negatively associated random variables~\cite[Prop. 7]{dubhashi1998}.

@@

\begin{definition}[Negative Association]
A set of random variables $X_1,..,X_n: \Omega \rightarrow \mathbb R$ is negatively associated if for all non-decreasing functions $f,g: \mathbb R^n \rightarrow \mathbb R$, which depend on disjoint sets of the variables, i.e., $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$ then
\[
\expect [f(X_1,\ldots,X_n) g(X_1,\ldots,X_n)] \leq \expect [f(X_1,\ldots,X_n)] \expect [g(X_1,\ldots,X_n)] \textrm{.}
\]
\end{definition}

@@

The resulting algorithm is depicted in \cref{alg:cvm_new}.
\begin{algorithm}[t]
	\caption{New unbiased and total CVM algorithm.}\label{alg:cvm_new}
	\begin{algorithmic}[1]
  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ s.t. $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{3l}{\delta})} }$, $\frac{1}{2} \leq f < 1$, s.t., $nf$ integer
  \For{$i \gets 1$ to $l$}
    \State $b \getsr \Ber(p)$ \Comment insert $a_i$ with probability $p$ (and remove it otherwise)
    \If{$b$}
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$}
      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment Select a random $nf$ subset of $\chi$
      \State $p \gets pf$
    \EndIf
  \EndFor
  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
  \end{algorithmic}
\end{algorithm}%
Note that it is possible to just choose $f = \frac{n-1}{n}$, i.e., discarding a random element from $\chi$ in the subsampling step.
The key advantage is that there is no line 11 and the algorithm never returns $\bot$.
And based on our discussion from the previous section it is also an unbiased algorithm, i.e., the expectation of the result is exactly $|A|$.

>>>>

\section{Negative Dependence and an Unbiased CVM Algorithm\label{sec:negdep}}
Sometimes, sets of random variables possess a property called negative association.
The concept is a generalization of independence.
It has similar closure properties as independent random variables.
More interestingly, the standard Chernoff-Hoeffding bounds still apply to negatively associated random variables~\cite[Prop. 7]{dubhashi1998}.

@@

\begin{definition}[Negative Association]\label{def:neg_assoc}
A set of random variables $X_1,..,X_n: \Omega \rightarrow \mathbb R$ is negatively associated if for all non-decreasing functions $f,g: \mathbb R^n \rightarrow \mathbb R$, which depend on disjoint sets of the variables, i.e., $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$ then
\[
\expect [f(X_1,\ldots,X_n) g(X_1,\ldots,X_n)] \leq \expect [f(X_1,\ldots,X_n)] \expect [g(X_1,\ldots,X_n)] \textrm{.}
\]
\end{definition}

@@

The resulting algorithm is depicted in \cref{alg:cvm_new}.
\begin{algorithm}[h]
	\caption{New unbiased and total CVM algorithm.}\label{alg:cvm_new}
	\begin{algorithmic}[1]
  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ s.t. $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{3l}{\delta})} }$, $\frac{1}{2} \leq f < 1$, s.t., $nf$ integer
  \For{$i \gets 1$ to $l$}
    \State $b \getsr \Ber(p)$ \Comment insert $a_i$ with probability $p$ (and remove it otherwise)
    \If{$b$}
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$}
      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment select a random $nf$ subset of $\chi$
      \State $p \gets pf$
    \EndIf
  \EndFor
  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
  \end{algorithmic}
\end{algorithm}%
Note that it is possible to just choose $f = \frac{n-1}{n}$, i.e., discarding a random element from $\chi$ in the subsampling step.
The key advantage is that there is no line 11 and the algorithm never returns $\bot$.
And based on our discussion from the previous section it is also an unbiased algorithm, i.e., the expectation of the result is exactly $|A|$.

### 2025-02-07 paper/sec/semra.tex
Paper:   * Add section on formalization of negatively dependend variables.   * Minor edits.

\cref{eq:integral_bind} is crucially used to establish invariants of the form introduced in~\cref{sec:intro}.
\end{description}

\todo{I think we should briefly say that a streaming algorithm is modeled as folding over the monad?}

>>>>

\cref{eq:integral_bind} is crucially used to establish invariants of the form we will use in~\cref{sec:invariants}.
\end{description}
%\todo{I think we should briefly say that a streaming algorithm is modeled as folding over the monad?}
